Over Twenty Years of Virtual Machine Development and Debugging Through Simulation

### 2018-08-14 paper.tex
pass over the text

\openSmalltalkVM was originally written in Smalltalk to specify and example the virtual machine (VM) implementation. The same code base was changed a little bit to allow Smalltalk-to-C compilation, generating effectively the production VM. Two execution models are effectively available, the simulation mode, executing the Smalltalk code on top of a Smalltalk VM, and the production mode, compiling the code to executable through C. Simulation is used to develop and debug the VM. Production is used to release the VM.

As the VM evolved, by introducing better garbage collector algorithms or a just-in-time compiler, both execution models co-evolved. For example, the simulation mode was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We discuss some of the limitations and how we dealt with it. Then, we focus on two specific use-cases, a bug in the just-in-time compiler and the development of a new full garbage collector compaction algorithm, and we show how the simulation infrastructure helped us. Lastly, we discuss how we use the simulation mode to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.

@@

To specify the language and explain how to write the virtual machine (VM) for it, the Smalltalk-80 crew wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample}. The VM was ported from the original specifications however part of the code base was narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}. Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled to native code to produce the production VM. At this point, the VM consisted mainly in an interpreter, a memory manager with a generational garbage collector (GC) and BitBlt, an extension for the user interface bit-based engine. These tree parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes (Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}). As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM. The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT), \emph{Cogit}, a template-based JIT. The existing simulator was not able to interpreter the machine code generated by the JIT and had to be extended by binding various processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).

In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. Section \ref{sec:GCExp} reports our experience developing the full GC with our infrastructure. Section \ref{sec:JITExp} explains how we fixed a bug in the Just-in-Time compiler using single stepping in machine code in the simulator. Section \ref{sec:Analysis} shows how we abuse the simulator to analyse the runtime and direct our performance decisions. Lastly, we discuss some related work and conclude.

@@

As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C. The Core VM code, also known as the object engine, is written in Slang and includes mainly the interpreter, the template-based JIT and the memory manager. The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C. 
%EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.

@@

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code to C code in a few files. This first compilation pass takes a few seconds. Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable. This second compilation pass may take several dozens of seconds the first time, then, it depends on what the programmer changed and what the C compiler can cache or not, but it usually also takes a few seconds to recompile the C code. The VM can be compiled in two main flavors, interpreter-only or interpreter+JIT. Although the version with the JIT is used in production, the interpreter version is convenient for special purposes, for example to debug the garbage collector or to evaluate new language features without dealing with the JIT complexity.

There are multiple reasons why the object engine is written in Slang and not in C. One of the reason is that the Slang-to-C compiler, using différents annotations, generates C code slightly different from the Slang code: for example, it duplicates the implementation of specific methods with specific constant operands in Slang to generate more efficient code in the interpreter. The main reason is VM simulation. By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution.

@@

in this section, we first describe briefly the start-up sequence of \openSmalltalkVM using snapshots. Then, we detail the simulation of the interpreter-only VM. The simulation of the full VM is a superset and is explained in the last subsection.

\paragraph{Start-up from snapshots.} 
Smalltalk is a snapshot-based language. A Smalltalk program is started from a snapshot, \emph{i.e.,} a memory dump of all live objects at a given point in the execution of a program. The snapshot includes objects such as the classes, the complied methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory and resumes execution in the active process at snapshot time.

When programming with Smalltalk, the programmers usually start ups from a snapshot which contains the core libraries, the development environment and the application developed. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the snapshots. When the changes are done, the developer takes a new snapshot, which includes its changes. Deployment is done from a snapshot containing the deployed application and required core librairies (unused librairies and development tools may be removed from the deployed snapshot).

Starting up the VM for source files has never been possible. However, recent work~\cite{PharoBootstrap} allowed one of the Smalltalk runtime to recreate a snapshot from sources, indirectly allowing to start the VM from source files.

@@

The heap is simulated as a large contiguous byte array. References between objects are effectively indexes inside the byte array instead of pointers. All the C variables are simulated as Smalltalk objects. They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs for example). One of the most complex component is the Stack. The Stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. Each stack page is represented as a Smalltalk object.

All the Slang code is implemented in multiple Smalltalk classes, to organise the code. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. No polymorphism is available at runtime. However, polymorphism can be abused for debugging purposes. For example, a class is available holding all the garbage collection compaction logic. This class might be subclassed with a simulation only version: this version can use normal Smalltalk code over Slang, allowing to express easily specific constraints which leads to assertion failures if not met, such constraints are convenient for debugging.

@@

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory now holds the machine code generated and installed by the JIT. Bindings to processor simulator libraires (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated. Calls in-between slang code and direct machine code are a little bit trickier to simulate. Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator. This trap is caught is the VM simulator, which then resumes Slang simulation by calling the method corresponding to the invalid processor instruction. Calling machine code from Slang requires to start the simulation and then launch an exception to stop Slang simulation. The processor simulator can be start in two different ways. It can either start simulating code until it meets an invalid instruction or simulate one instruction at a time. The second version is slower, but allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the template JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle. To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which is started on the compiled method and only includes the subset of objects required by the JIT to generate the machine code.

>>>>

\openSmalltalkVM was originally written in Smalltalk to specify and example the virtual machine (VM) implementation. The same code base was changed a little bit to allow Smalltalk-to-C compilation to generate the production VM through the C compiler. Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler. Simulation is used to develop and debug the VM. Production is used to release the VM.

As the VM evolved, by introducing better garbage collector algorithms or a just-in-time compiler, both execution models co-evolved. For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.

In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it. We mention some of the limitations and how we worked around them. We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.

@@

To specify the language and explain how to write the virtual machine (VM) for it, the Smalltalk-80 crew wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample}. The VM was ported from the original specifications. Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture} and generate an efficient VM through the C compiler. Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM. At this point, the VM consisted mainly in:
\begin{itemize}
\item an interpreter with a spaghetti stack, 
\item a memory manager with a generational garbage collector (GC) and
\item BitBlt, an extension for the user interface bit-based engine.
\end{itemize} 
These three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}. As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM. The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). The existing simulator was not able to simulate the machine code generated by the JIT and had to be extended by binding multiple processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).

In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. Section \ref{sec:GCExp} and Section \ref{sec:JITExp}  reports our experience developing respectively the full GC and the JIT with our infrastructure. Section \ref{sec:Analysis} shows how we abuse the simulator to analyse the runtime and direct design decisions to improve performance. Lastly, we discuss some related work and conclude.

@@

As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C. The Core VM code (the interpreter, the JIT and the memory manager), is written in Slang. The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C. 
%EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.

@@

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few files. This first step takes several seconds. Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable. This second step may take several dozens of seconds the first time, then, depending on what the programmer changed and what the C compiler have cached, it usually only takes a few seconds to recompile the C code.

The VM can be compiled in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is used in production, the interpreter version is convenient for special purposes. For example, debugging the garbage collector or evaluating new language features without dealing can be done in the interpreter-only VM, avoiding the JIT complexity.

There are multiple reasons why we keep writing the core components in Slang and not in C/C++. Most of them are small details. For instance, the Slang-to-C compiler generates C code slightly different from the Slang code using different annotations. It can for example duplicate the implementation of specific methods with specific constant operands to generate more efficient code in the interpreter. The main reason why we keep using Slang over C/C++ is VM simulation. By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing to develop and debug it with the Smalltalk development tools and a live programming experience.

@@

In this section, we firstly describe briefly the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success. Then, we detail the simulation of the interpreter-only VM. The simulation of the full VM is a superset and is explained in the last subsection.

\paragraph{Start-up from snapshots.} 
Smalltalk is a snapshot-based language. A Smalltalk program is started from a snapshot, \emph{i.e.,} a memory dump of all live objects at a given point in the execution of a program. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory and resumes execution in the process active at snapshot time.

When programming with Smalltalk, the programmers usually starts from a snapshot which contains the core libraries, the development environment and the application developed. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the set of live objects. When the changes are done, the developer takes a new snapshot, which includes its changes. Deployment is performed using a snapshot containing the deployed application and the required core librairies (unused librairies and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).

Starting up \openSmalltalkVM from source files has never been possible. However, recent work~\cite{PharoBootstrap} allowed one of the Smalltalk runtime to recreate a snapshot from sources, indirectly allowing to start the VM from source files.

@@

The heap is simulated as a large contiguous byte array. References between objects are effectively indexes inside the byte array instead of pointers. All the C variables are simulated as Smalltalk objects. They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). An interesting example is the execution stack. The stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM. Each stack page is represented as a Smalltalk object.

All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism. For example, the AbstractCompactor class has two subclasses, one implements a sweep algorithm and the other a compact one. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. All the flexibility is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be abused for debugging purposes. Still re-using the same example, the AbstractCompactor class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.

@@

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraires (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated. Calls in-between slang code and direct machine code are a little bit trickier to simulate. Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator. Such traps can be caught is the VM simulator, which then resumes Slang simulation by calling the correct method based on the invalid processor instruction. Calling machine code from Slang requires to start the machine code simulation but also to raise an exception to stop Slang simulation. Indeed, in the production VM, the processor can execute the C code or the native code generated by the JIT, but not both at the same time.

Machine code simulation can be performed in two different ways. The processor simulator can start simulating code until it meets an invalid instruction. That version is convenient because it is the fastest to execute, while still catching errors such as invalid memory accesses. Alternatively, it can simulate one instruction at a time. This second version is slower, but it allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid. To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}. HERE Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which is started on the compiled method and only includes the subset of objects required by the JIT to generate the machine code.
